---
title: "Bootstrap Simulations"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Introduction

Simulate a toy dataset with a *specified* signal-to-noise ratio (SNR), allowing the predictors $x_1$ and $x_2$ to contribute UNEQUALLY to the signal.

Definition used here (Hastie, Tibshirani & Friedman, ESL Sec. 16.2.2, adapted to a Bernoulli outcome):

$SNR = Var( \eta(X) ) / E[ Var(Y | \eta(X)) ]$

where $eta(X) = \beta_0 + \beta_1*x_1 + \beta_2*x_2$ is the linear predictor (log-odds), $Var(\eta(X))$ is the variance of the log-odds across the population ("signal"), and $E[Var(Y|\eta(X))] = E[p (1-p)]$ is the average Bernoulli variance implied by the model ("noise").

A LARGER SNR means $x_1$ and/or $x_2$ explain more of the variation in $y$ relative to irreducible Bernoulli randomness, which also pushes the c-statistic (AUC) higher. The two are linked but not identical, so we calibrate on SNR directly and report the resulting AUC as a by-product.

Instead of forcing the coefs of $x_1$ and $x_2$ to $\beta_1 = \beta_2$ (equal contribution), we fix a *relative contribution ratio* between $x_1$ and $x_2$, then solve for an overall scale factor that hits the target SNR while preserving that ratio.

```{r}
#| label: init-setup

library(pROC)
library(DBI)
library(duckdb)

set.seed(2026)
n <- 200000
prevalence <- 0.20     # target proportion of events for y
```

## Step-by-step procedure

```{r}
#| label: Step-by-step-procedure

## Step 1: Simulate the predictor variables

x1 <- rnorm(n, mean = 0, sd = 1)   # causal predictor 1
x2 <- rnorm(n, mean = 0, sd = 1)   # causal predictor 2
x3 <- rnorm(n, mean = 0, sd = 1)   # noise variable, unrelated to y, x1, x2


## Step 2: Specify how much MORE x1 should contribute than x2
# ratio = 1   -> x1 and x2 contribute equally
# ratio = 2   -> x1's coefficient is twice x2's, so x1 contributes 4x the
#                *variance* of x2 (since variance scales with beta^2),
#                i.e. x1 supplies 80% of the total signal variance and
#                x2 supplies 20%.
# ratio < 1   -> x2 dominates instead of x1.
contribution_ratio <- 2.0   # beta1 / beta2; edit this to change the split

## Step 3: Function to compute achieved SNR for a given overall "scale", holding the beta1:beta2 ratio fixed at contribution_ratio

# beta1 = contribution_ratio * scale
# beta2 = scale
# Scaling both coefficients together by "scale" changes the overall signal strength (and hence SNR) without changing the relative split between x1 and x2.
get_snr_unequal <- function(scale, ratio, x1, x2, prevalence) {

  beta1 <- ratio * scale
  beta2 <- scale
  lp_signal <- beta1 * x1 + beta2 * x2
  
  # Inner function: for a candidate intercept b0, compute how far the
  # implied average probability is from the target prevalence.
  prevalence_gap <- function(b0) {
    eta <- b0 + lp_signal
    p   <- 1 / (1 + exp(-eta))
    mean(p) - prevalence
  }
  
  # Solve for the intercept b0 that makes the average predicted
  # probability equal to the target prevalence (root-finding).
  b0 <- uniroot(prevalence_gap, lower = -10, upper = 10)$root

  eta <- b0 + lp_signal
  p   <- 1 / (1 + exp(-eta))

  var_signal <- var(eta)            # variance of the log-odds (signal)
  var_noise  <- mean(p * (1 - p))   # average Bernoulli variance (noise)

  list(snr = var_signal / var_noise, beta0 = b0, beta1 = beta1, beta2 = beta2,
       eta = eta, p = p)
}


## Step 4: Solve for the "scale" that achieves the target SNR, given the fixed contribution_ratio between beta1 and beta2

# Change target_snr to whatever value you want to simulate.
# As a rough guide (derived from this same predictor setup):
#   SNR ~ 1   -> weak signal,     AUC ~ 0.60
#   SNR ~ 4   -> low-moderate,    AUC ~ 0.69
#   SNR ~ 8   -> moderate signal, AUC ~ 0.75
#   SNR ~ 16  -> strong signal,   AUC ~ 0.81
target_snr <- 8   # moderate signal

snr_gap <- function(scale) {
  get_snr_unequal(scale, contribution_ratio, x1, x2, prevalence)$snr - target_snr
}

scale_solution <- uniroot(snr_gap, lower = 0.001, upper = 5)$root

# Recompute the final quantities at the solved scale
result <- get_snr_unequal(scale_solution, contribution_ratio, x1, x2, prevalence)
beta0 <- result$beta0
beta1 <- result$beta1
beta2 <- result$beta2
eta   <- result$eta
p_y   <- result$p

cat("Solved beta1 (x1 coefficient):", round(beta1, 4), "\n")
cat("Solved beta2 (x2 coefficient):", round(beta2, 4), "\n")
cat("Solved intercept (beta0):", round(beta0, 4), "\n")
cat("Achieved SNR:", round(result$snr, 3), "(target was", target_snr, ")\n")

# Report how the total signal variance splits between x1 and x2 -
# this is the practical, interpretable readout of "who contributes more."
var_x1_contribution <- var(beta1 * x1)
var_x2_contribution <- var(beta2 * x2)
total_contribution  <- var_x1_contribution + var_x2_contribution
cat("Share of signal variance from x1:",
    round(var_x1_contribution / total_contribution, 3), "\n")
cat("Share of signal variance from x2:",
    round(var_x2_contribution / total_contribution, 3), "\n")


## Step 5: Draw y, assemble the dataset

y <- rbinom(n, size = 1, prob = p_y)
cat("Empirical event rate:", round(mean(y), 3), "\n")

sim_data <- data.frame(y = y, x1 = x1, x2 = x2, x3 = x3)


## Step 6: Fit the logistic model and confirm x1's larger coefficient
## shows up as a larger effect size, plus check overall AUC

fit <- glm(y ~ x1 + x2, data = sim_data, family = binomial())
summary(fit)   # beta1 estimate should be roughly contribution_ratio times beta2

pred_probs <- predict(fit, type = "response")
roc_obj <- roc(sim_data$y, pred_probs, quiet = TRUE)
cat("C-statistic (AUC):", round(auc(roc_obj), 3), "\n")


## Step 7 (sanity check): x3 remains pure noise, unaffected by the SNR
## setting on x1/x2, since it never enters the data-generating process
noise_check <- glm(y ~ x3, data = sim_data, family = binomial())
summary(noise_check)
```

## Reusable function version

```{r}
## Parameterized by target SNR AND the relative contribution ratio between x1 and x2

simulate_snr_dataset_unequal <- function(target_snr, ratio = 1, n = 200000,
                                          prevalence = 0.2, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)

  x1 <- rnorm(n)
  x2 <- rnorm(n)
  x3 <- rnorm(n)

  snr_gap <- function(scale) get_snr_unequal(scale, ratio, x1, x2, prevalence)$snr - target_snr
  scale_sol <- uniroot(snr_gap, lower = 0.001, upper = 5)$root
  res <- get_snr_unequal(scale_sol, ratio, x1, x2, prevalence)

  y <- rbinom(n, size = 1, prob = res$p)

  data.frame(y = y, x1 = x1, x2 = x2, x3 = x3,
             beta1 = res$beta1, beta2 = res$beta2)  # coefficients attached for reference
}
```

## Example datasets

```{r}
# Example: moderate SNR (8), x1 twice as influential as x2
dat_unequal <- simulate_snr_dataset_unequal(target_snr = 8, ratio = 2, seed = 1)

# Example: moderate SNR (8), equal contribution (ratio = 1)
dat_equal <- simulate_snr_dataset_unequal(target_snr = 8, ratio = 1, seed = 1)
```

## Saving the datasets

```{r}
# calculations.qmd
con <- dbConnect(duckdb(), dbdir = "data/sim_data.duckdb")

dbWriteTable(con, "dat_unequal", dat_unequal, overwrite = TRUE)

dbWriteTable(con, "dat_equal", dat_equal, overwrite = TRUE)

# A check that the tables are in the DB:
dbListTables(con)

# Disconnect:
dbDisconnect(con, shutdown = TRUE)
```
